---
title: Visualizing Pearl Jam's songs
author: Jordan Upton
date: '2019-10-08'
slug: visualizing-pearl-jam-s-songs
categories:
  - R
tags:
  - plot
  - spotify
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I'm a big Pearl Jam fan, though that term is a bit of a [rabbit hole](https://xkcd.com/1095/).  Surely, I thought, there must be a way to combine my love of Pearl Jam with my love of data visualisation.  Surely, I have nothing better to do!

Fortunately, Spotify has an [open API](https://developer.spotify.com/my-applications/#!/applications) that allows users to extract data not just on artists and songs but also provides some proprietary metrics such as 'danceability', 'energy' and 'loudness'.  Even better, there's already [an R package](https://cran.r-project.org/web/packages/spotifyr/spotifyr.pdf) that makes it easy to work with the API.

I pulled Spotify's metrics for the 134 songs in Pearl Jam's studio catalog and ran them through a [principal component analysis](https://en.wikipedia.org/wiki/Principal_component_analysis) (PCA).  PCA allows us to take a large number of metrics and "compress" them into a smaller number of metrics which retain much of the same information as the original.  It's sort of like a shadow of our data; not exactly like the original, but still enough for us to get an idea of what it looks like.

I used the first two resulting principal components (which accounted for 59% of the total variance) to generate a scatterplot of the songs.  Each song was then color-coded by album, using the average color that appeared on that album's cover art.

```{r, echo = FALSE, dpi=96*3, fig.align='center', fig.width=16, fig.height=9, out.width="100%", message=FALSE, warning = FALSE}
library(tidyverse)
library(lubridate)
library(ggthemes)
library(scales)
library(broom)
library(ggrepel)
library(spotifyr)
library(jpeg)
library(grid)
library(gridExtra)

#grab a custom function for captioning plots
source("../../static/functions/functions.R")


#First, you'll need a Spotify developer account.  Sign up here
#https://developer.spotify.com/my-applications/#!/applications
#Then, input your client id and secret below
#Sys.setenv(SPOTIFY_CLIENT_ID = 'x')
#Sys.setenv(SPOTIFY_CLIENT_SECRET = 'x')

#get an access token
access_token <- get_spotify_access_token()

#find the artist id for Pearl Jam
artist_id <- search_spotify("Pearl Jam", "artist", authorization = access_token) %>%
  top_n(1, popularity) %>%
  pull(id)

#get all of the albums for the artist
albums <- get_artist_albums(artist_id, authorization = access_token) %>%
  #do some cleanup to limit to the official albums
  filter(name %in% c("Ten", "Vitalogy", "Vs.", "No Code", "Yield", "Binaural",
                     "Riot Act", "Pearl Jam", "Backspacer", "Lightning Bolt"),
         release_date != ymd(20090101))


albums$color <- NA
#for each album, download the album art and extract the mean color
for(i in seq_len(nrow(albums))){

  tf <- tempfile(fileext = ".jpg")
  download.file(albums$images[[i]]$url[1], destfile = tf, mode = "wb")
  art <- readJPEG(tf)

  #the art file gets stored as an x by y by 3 array, with one dimension each for RGB.
  #By taking the mean value of each RGB dimension and combining them, we get the average color
  albums$color[i] <- rgb(mean(art[,,1]),mean(art[,,2]),mean(art[,,3]))
}
rm(tf, art)


#create a tibble to store data for all tracks on these albums
tracks <- tibble()

#loop through the albums and get all track data
for(i in seq_len(nrow(albums))){
  tracks <- tracks %>%
    bind_rows({
      get_album_tracks(albums$id[i], authorization = access_token) %>%
        mutate(album_id = albums$id[i])
    })
}


#create a tibble to store data for all the features of each track
features <- tibble()

#loop through the tracks and get their features
for(i in seq_len(nrow(tracks))){
  features <- features %>%
    bind_rows(get_track_audio_features(tracks$id[i], authorization = access_token))
}

#join the desireable features onto the track data
tracks <- tracks %>%
  left_join({
    features %>%
      select(id, danceability, energy, loudness, speechiness, acousticness,
             instrumentalness, liveness, valence)
  }, by = "id")



#conduct a principal components analysis of the track features and extract the first 2 PCs
tracks <- tracks %>%
  bind_cols({
    tracks %>%
      select(danceability, energy, loudness, speechiness, acousticness,
             instrumentalness, liveness, valence) %>%
      prcomp(center = TRUE, scale = TRUE) %>%
      tidy("x") %>%
      spread(key = PC, value = value) %>%
      select(PC1 = `1`, PC2 = `2`)
  }) %>%
  #do a little cleanup on track names
  mutate(name2 = str_remove(name, " \\(Remastered\\)| - Remastered| - bonus track")) %>%
  #join on the album color
  left_join(albums %>% select(id, color), by = c("album_id" = "id"))


#set the random seed to ensure reproducibility of the plot
set.seed(20191008)

#plot the tracks by their first two principal components, colored by the album
p <- tracks %>% 
  ggplot(aes(PC1, PC2)) + 
  geom_label_repel(aes(label = name2, fill = color),
                   segment.color = NA, color = "grey94", fontface = "bold", size = 3) + 
  #need to let ggplot know we're giving it the color hex codes directly
  scale_fill_identity() + 
  #make some general changes to the theme to make everything look nice
  theme_fivethirtyeight() + 
  theme(axis.title = element_blank(),
        axis.text = element_blank(),
        legend.position = "none",
        plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"),
        panel.grid = element_blank()) + 
  labs(title = "\"I've figured out numbers and what they're for\"",
       subtitle = "The songs of Pearl Jam")

#grab the range of the plot so we can automatically annotate based on them
plot_range <- tibble(x = ggplot_build(p)$layout$panel_scales_x[[1]]$range$range,
                     y = ggplot_build(p)$layout$panel_scales_y[[1]]$range$range)

p <- p +
  #add some explainer text
  annotate("text", 
           x = min(plot_range$x) + 0.3 * abs(diff(plot_range$x)), 
           y = min(plot_range$y) + 0.05 * abs(diff(plot_range$y)),
           label = "Distance between songs is a rough measure of their similarity.\nX and Y position determined using the first two principal components from an analysis on song\ndanceability, energy, loudness, speechiness, acousticness, instrumentalness, liveness & valence.\nLabels are colored by album, using the average color of the album art.",
           color = "grey40", size = 4, fontface = "bold", family = "sans") 



#caption the plot
caption_plot(p, "DatArt.netlify.com", "Source: Spotify") %>% 
  grid.draw()
```

The distance between the songs is a rough measure of how similar they are, at least in terms of Spotify's metrics.  For example, the far left contains a lot of softer, slower tunes.  To the top and right are some of the group's more pop-rock numbers while the bottom right contains some real hard, sometimes depressing stuff.  It's interesting to see how clustered most of the songs from 'Ten' are, and how much they've expanded their sound since then.

If nothing else, this confirms my long-held suspicion that 'Push Me, Pull Me' is just weird.